### 2013-01-07 ISMB/methods_ISMB.tex
Method v2

We show how a modification of the dynamic programming scheme introduced in the previously introduced \RNAmutants allow for the sampling of good and diverse design candidates in linear time and space complexities.

@@

It was established~\cite{Waldispuhl2011} that, for each value of $x$, there exists constants $\mu_x$ and $\sigma_x$ such that the \GCContent distribution observed in sequences sampled with respect to the above distribution converges to a normal law of expectation in $\mu\cdot n\cdot(1+o(1))$ and standard deviation in $\sigma\cdot\sqrt{n}\cdot(1+o(1))$.
Furthermore, it can easily be shown that the expected \GCContent is a continuous and strictly increasing monotonic  function of $x$, whose limits are $0$ when $x=0$ and $n$ when $x\to +\infty$.

Consequently, for any targeted \GCContent $gc\in[0\%,100\%]$, there exists a unique value $x_{gc}$ such that generated sequences feature, on the average, the right \GCContent. Such a value can be determined using any numerical recipe, and a simple bisection is used in our implementation. Furthermore, the concentration of the distribution, as asserted by the limited standard deviation, implies a growth in $\Theta(\sqrt{n})$ (resp. $\Theta(1)$, i.e. constant) for the expected number of attempts before generating a sequence having the exact targeted \GCContent  (resp. within $[gc-\kappa,gc+\kappa]$, where $\kappa\in\Theta(1/\sqrt n)$).

@@

Let us now describe a linear-time algorithm to sample sequences at random in the pseudo-Boltzmann distribution. This algorithm follows the general principles of the recursive approach to random generation~\cite{Wilf1977}, pioneered in the context of RNA by the \SFold algorithm~\cite{Ding2003}. The algorithm starts by precomputing the partition function restricted to each valid interval, and then performs a series of stochastic backtracks, using precomputed values to decide on the probability of each alternative.

\subsubsection{Precomputation}
Let us describe a dynamic programming algorithm to compute $\Z{i,j}{a,b}$, the pseudo-partition function restricted to an interval $[i,j]$, assuming its (previously chosen) flanking nucleotides are $a$ and $b$ respectively.

Since the only sequence over an empty interval is the empty sequence, having energy $0$, one has
\begin{equation}
	\forall i \in [0,n-1]:\, \Z{i+1,i}{a,b}=1.
	\label{eq:Z_in}
\end{equation}

The recursion consists in four terms, depending on the base-pairing status and context of $(i,j)$:
\begin{equation*}
	\Z{i,j}{a,b}:=\left\{
  \begin{array}{ll}
  		\displaystyle
      \sum_{a'\in \B}  
      x^{\gc(a')}
      \cdot\Z{i+1,j}{a',b} &\begin{array}{@{}l@{}}\text{If }\Target_{i}=-1\\ \text{\relsize{-1}Position $i$ unpaired,}\end{array}\\
      \displaystyle
      \sum_{a',b'\in \B^2}
			 x^{\gc(a'.b')}
			 \cdot e^{\frac{-\ES^{\beta}_{ab \to a'b'}}{RT}}
			 \cdot \Z{i+1,j-1}{a',b'}&
\begin{array}{@{}l@{}}\text{Elif }\Target_i=j \land \Target_{i-1}=j+1\\ \text{\relsize{-1}$(i,j)$ base-paired, stacked onto $(i$-$1,j$+$1)$,}\end{array}
			 \\
			 \displaystyle
      \sum_{a',b'\in \B^2}
      x^{\gc(a'.b')}
			\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
      \cdot\Z{i+1,k-1}{a',b'}
      \cdot\Z{k+1,j}{b',b} & 
\begin{array}{@{}l@{}}\text{Elif }\Target_i=k \land i < k \leq j\\
\text{\relsize{-1} Junction or bulge $\Rightarrow$ split, no stacking.}
\end{array}
\\
      0 &\text{Otherwise}
	\end{array}\right.
\label{eq:Z_rec}
\end{equation*}
It is worth mentioning that, despite the appearances, this recurrence can be computed in linear $\Theta(|\Target|)$ time and space. Indeed, one can show that position $j$ is entirely determined by $i$ within subsequent recursive calls.

More precisely, let us consider the following predicate: {\em Let $(i',j')$ be the base-pair such that $i'<i<j'$ while minimizing $i-i'$, then TODO!!!!

$j$ is the position preceding the end of the closest base-pair nesting position $i$ if such a base-pair exists, or the last position in the sequence otherwise.} Obviously, the predicate holds for $[0,n-1]$. Moreover, if $i$ is unpaired, then $i+1$ and $i$

%The cases can be broken down as follows:
%\begin{description}
%\item[$S_{i}=-1$:] If the nucleotide at position $i$ is unpaired, then 
%any sequence consists in a nucleotide $a'$ at position  $i$, 
%followed by a sequence over $[i+1,j]$ and having flanking nucleotides $a'$ and $b$.
%\item[$S_i=j$ and $S_{i-1}=j+1$:] Any sequence generated in $[i,j]$ consists of two nucleotides $a'$ and $b'$, flanking a sequence over $[i+1,j-1]$.
%Since positions $i$ and $i-1$ are paired with $j$ and $j+1$ respectively, 
%then a stacking energy contribution is added. 
%\item[$S_i=k$ and $i<k \leq j$:] If position $i$ is paired and not involved in a stacking, then the 
%only term contributing directly to the energy is the nucleotide distribution $\Prob(\cdot)$. 
%Any sequence on $[i,j]$ consists of two nucleotide $a'$ and $b'$ at positions $i$ and $k$ respectively, flanking a sequence over interval $[i+1,k-1]$ and preceding a (possibly empty) sequence interval $[k+1,j]$.
%\item[Else:] In any other case, we are in a derivation of the SCFG that does not correspond to the secondary structure $S$, and we return $0$.
%\end{description}

\subsubsection{Stochastic backtrack}
Once the partition functions have been computed, a stochastic backtrack starts from the complete interval $[0,|\Target|-1]$ and, at each step, chooses a suitable assignment for one or several positions, as described in Algorithm~\ref{alg:back}. 
\begin{algorithm}[t]
	\SetAlgoLined
\SetKwFunction{Backtrack}{Backtrack}
\SetKwFunction{Random}{Random}
	rand $\leftarrow$ \Random$\left(\Z{i,j}{a,b}\right)$\;
	k $\leftarrow S_i$\;
	\If(\tcp*[f]{Unpaired}){$k = -1$}{
		\For{$a'\in\B$}{
			rand $\leftarrow$ rand $- x^{\gc(a')}\cdot \Z{i+1,j}{a',b}$\;
			\lIf{$\text{rand}<0$}\Return{$a'.\Backtrack\left((i+1,j),(a',b),S\right)$}\;
		}
	}
	\Else{
		\If(\tcp*[f]{Stacking pairs}){$k=j \land S_{i-1}=j+1$}{
			\For{$(a',b')\in\B\times\B$}{
				rand $\leftarrow$ rand $-
			 	x^{\gc(a'.b')}
			 	\cdot e^{\frac{-\ES^{\beta}_{ab \to a'b'}}{RT}}
			 	\cdot \Z{i+1,j-1}{a',b'}	$\;
				\If{$\text{rand}<0$}{
					\Return{$a'.\Backtrack\left((i+1,j-1),(a',b'),S\right). b'$}\;		
				}
			}
		}
		\Else{
			\If(\tcp*[f]{Paired, not stacked}){$\Target_i=k \land i < k \leq j$}{
				\For{$(a',b')\in\B\times\B$}{
					rand $\leftarrow$ rand $-	
       x^{\gc(a'.b')}
					\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
  	    	\cdot\Z{i+1,k-1}{a',b'}
	    	  \cdot\Z{k+1,j}{b',b}$\;	
 					\If{$\text{rand}<0$}{
						\Return{$a'
						.\Backtrack\left((i+1,k-1),(a',b'),S\right)
						.b'
						.\Backtrack\left((k+1,j),(b',b),S\right)
						$}\;	
					}
				}
		 }
    \Else {\Return{$\varepsilon$};	\tcp*[f]{Empty sequence}}
		}
	}
\caption{\protect\Backtrack$\left((i,j),(a,b),S\right)$\label{alg:back}}
\end{algorithm}

\subsection{Overall complexity}

>>>>

We show how a modification of the dynamic programming scheme introduced in the previously contributed\RNAmutants allows for the sampling of good and diverse design candidates, in linear time and space complexities.

@@

@@

Let us now describe a linear-time algorithm to sample sequences at random in the pseudo-Boltzmann distribution. This algorithm follows the general principles of the recursive approach to random generation~\cite{Wilf1977}, pioneered in the context of RNA by the \SFold algorithm~\cite{Ding2003}. The algorithm starts by precomputing the partition function restricted to each valid interval, and then performs a series of recursive stochastic backtracks, using precomputed values to decide on the probability of each alternative.

\subsubsection{Precomputing the pseudo-partition function}\label{sec:pf}
Firstly, a dynamic programming algorithm computes $\mathcal{Z}\substack{(i,j)\\ [a,b]}$ the pseudo-partition function restricted to an interval $[i,j]$, assuming its (previously chosen) flanking nucleotides are $a$ and $b$ respectively. 
Since the only sequence over an empty interval is the empty sequence, having energy $0$, one has
\begin{equation}
	\forall i \in [0,n-1]:\, \Z{i+1,i}{a,b}=1.
	\label{eq:Z_in}
\end{equation}

The recursion consists in four terms, depending on the base-pairing status and context of $(i,j)$:
\begin{equation}
	\Z{i,j}{a,b}:=\left\{
  \begin{array}{ll}
  		\displaystyle
      \sum_{a'\in \B}  
      x^{\gc(a')}
      \cdot\Z{i+1,j}{a',b} &\begin{array}{@{}l@{}}\text{If }\Target_{i}=-1\\ \text{\relsize{-1}Position $i$ unpaired,}\end{array}\\
      \displaystyle
      \sum_{a',b'\in \B^2}
			 x^{\gc(a'.b')}
			 \cdot e^{\frac{-\ES^{\beta}_{ab \to a'b'}}{RT}}
			 \cdot \Z{i+1,j-1}{a',b'}&
\begin{array}{@{}l@{}}\text{Elif }\Target_i=j \land \Target_{i-1}=j+1\\ \text{\relsize{-1}$(i,j)$ stacking pair onto $(i$-$1,j$+$1)$,}\end{array}
			 \\
			 \displaystyle
      \sum_{a',b'\in \B^2}
      x^{\gc(a'.b')}
			\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
      \cdot\Z{i+1,k-1}{a',b'}
      \cdot\Z{k+1,j}{b',b} & 
\begin{array}{@{}l@{}}\text{Elif }\Target_i=k \land i < k \leq j\\
\text{\relsize{-1}Junction/bulge $\Rightarrow$ split, no stack,}
\end{array}
\\
      0 &\text{Otherwise.}
	\end{array}\right.
\label{eq:Z_rec}
\end{equation}

Despite its apparent quadratic complexity, the recurrence described in Subsection~\ref{sec:pf} can be computed in linear $\Theta(|\Target|)$ time and space, owing to the fact that, on any recursive call, the position $j$ is entirely determined by $i$.

\begin{theorem}Starting from the complete region $[0,n-1]$, Equation~\eqref{eq:Z_rec} can be computed in $\Theta(|\Target|)$ time and space complexity.
\end{theorem}
\begin{proof}

Consider the {\bf covering base-pair} $\CNBP{i}$ of a position $i$, defined as the base-pair $(i',j')$ such that $i'<i<j'$ and which minimizes $i-i'$. For the sake of simplicity, and without loss of generality, let us assume that there exists a virtual base-pair $(-1,n)$ in $\Target$, so that any position in $[0,n-1]$ has a covering base-pair.
Then one easily shows by induction that, starting from the complete interval $[0,n-1]$, the recursive calls only evaluate $\mathcal{Z}$ over intervals $[i,j]$ such that $j=j'-1$, $\CNBP{i}=[i',j']$.

\begin{cframed}This property clearly holds initially for the complete interval $[0,n-1]$, owing to the virtual base-pair $(-1,n)$. Furthermore, if it holds for a given $(i,j)$, $\CNBP{i}=[i',j+1]$, then:
\begin{itemize} 
  \item If $i$ is unpaired (Case 1), then $\CNBP{i+1}=\CNBP{i}=[i',j+1]$, and the property is satisfied by the subsequent call over $[i+1,j]$. 
  \item If $i$ is paired with $j$ and stacking onto $(i-1,j+1)$ (Case 2), then one has $\CNBP{i+1}=[i,j]$ and the  subsequent call over $[i+1,j-1]$ satisfies the property.
  \item If $i$ is paired with some $k$ (Case 3), then $\CNBP{i+1}=[i,k]$ and $\CNBP{k+1}=[i',j+1]$. Therefore the subsequent calls, respectively over $[i+1,k-1]$ and $[k+1,j]$, both satisfy the property.
\end{itemize}
The case of an empty sequence does not cause a recursive call and must not be verified.
\end{cframed}
Since, in any interval $[i,j]$ satisfying the property, the ending position $j$ is completely determined by $i$, then there exists at most $\Theta(|\Target|)$ intervals that need to be evaluated, and we conclude on the overall complexity with the remark that $a$, $b$, $a'$ and $b'$ vary over bounded ranges and only contribute to a constant in the complexity.
\end{proof}

\subsubsection{Stochastic backtrack}
Once the pseudo-partition functions have been computed, a stochastic backtrack starts from the complete interval $[0,|\Target|-1]$ and, at each step, chooses a suitable assignment for one or several positions, using probabilities derived from the precomputation. One or several recursive calls over regions that are left unassigned are then performed, as described by Algorithm~\ref{alg:back}.

Since, during each recursive call, the algorithm assigns at least one nucleotide to a -- previously unassigned -- position, and that the number of executions of the loops are bounded by a constant, then the complexity of the algorithm grows linearly on the targeted length. 
\begin{algorithm}[t]
\DontPrintSemicolon
	\SetAlgoLined
\SetKwFunction{Backtrack}{Backtrack}
\SetKwFunction{Random}{Random}
	rand $\leftarrow$ \Random$\left(\Z{i,j}{a,b}\right)$\tcp*[r]{Draw random number in $[0,\Z{i,j}{a,b}[$}
 \lIf(\tcp*[f]{Empty sequence}){$j<i$}{\Return{$\varepsilon$}}
	\Else{
  k $\leftarrow S_i$\;
	\If(\tcp*[f]{Unpaired}){$(k = -1)$}{
		\For{$a'\in\B$}{
			rand $\leftarrow$ rand $- x^{\gc(a')}\cdot \Z{i+1,j}{a',b}$\;
			\lIf{$\text{rand}<0$}\Return{$a'.\Backtrack\left(i+1,j,a',b,\Target\right)$}\;
		}
	}
	\lElse{
		\If(\tcp*[f]{Stacking pairs}){$(k=j) \land (S_{i-1}=j+1)$}{
			\For{$(a',b')\in\B\times\B$}{
				rand $\leftarrow$ rand $-
			 	x^{\gc(a'.b')}
			 	\cdot e^{\frac{-\ES^{\beta}_{ab \to a'b'}}{RT}}
			 	\cdot \Z{i+1,j-1}{a',b'}	$\;
				\lIf{$\text{rand}<0$}{
					\Return{$a'.\Backtrack\left(i+1,j-1,a',b',\Target\right). b'$}\;		
				}
			}
		}
		\lElse{
			\If(\tcp*[f]{Paired, not stacked}){$(\Target_i=k) \land (i < k \leq j)$}{
				\For{$(a',b')\in\B\times\B$}{
					rand $\leftarrow$ rand $-	
       x^{\gc(a'.b')}
					\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
  	    	\cdot\Z{i+1,k-1}{a',b'}
	    	  \cdot\Z{k+1,j}{b',b}$\;	
 					\If{$\text{rand}<0$}{
						\Return{$a'
						.\Backtrack\left(i+1,k-1,a',b',\Target\right)
						.b'
						.\Backtrack\left(k+1,j,b',b,\Target\right)
						$}\;	
					}
				}
		 }
		}
	}
 }
\caption{\protect\Backtrack$\left(i,j,a,b,\Target\right)$\label{alg:back}}
\end{algorithm}

\subsubsection{Rejecting unsuitable candidates}

Note that, although the parameter $x$ is introduced to match the expected \GCContent with the targeted one, nothing prevents Algorithm~\ref{alg:back} to generate sequences of arbitrary \GCContent. Therefore, we use a rejection-based approach~\cite{Bodini2010} previously used by the authors in a similar context~\cite{Waldispuhl2011}.

It was established~\cite{Waldispuhl2011} that, for each value of $x$, there exists constants $\mu_x$ and $\sigma_x$ such that the \GCContent distribution observed in sequences sampled with respect to the above distribution converges asymptotically towards a normal law having expectation in $\mu\cdot n\cdot(1+o(1))$ and standard deviation in $\sigma\cdot\sqrt{n}\cdot(1+o(1))$.
Furthermore, it can easily be shown that the expected \GCContent is a continuous and strictly increasing monotonic  function of $x$, whose limits are $0$ when $x=0$ and $n$ when $x\to +\infty$.

Consequently, for any targeted \GCContent $gc\in[0\%,100\%]$, there exists a unique value $x_{gc}$ such that generated sequences feature, on the average, the right \GCContent. Such a value can be determined using any numerical recipe, and a simple bisection procedure, further described in earlier work~\cite{Waldispuhl2011}, is used in our implementation. Furthermore, the concentration of the distribution, as asserted by the limited standard deviation, implies a growth in $\Theta(\sqrt{n})$ (resp. $\Theta(1)$, i.e. constant) for the expected number of attempts before generating a sequence having the exact targeted \GCContent  (resp. within $[gc-\kappa,gc+\kappa]$, where $\kappa\in\Theta(1/\sqrt n)$).

\subsection{Redesigning unpaired regions: A local/global hybrid approach}
